Fully Extrinsic Type Inference (DRAFT 2025-06-20)

### 2025-08-08 doc/extrinsic_type_inference_paper/paper.tex
update model typing definition in paper
update model typing lfp rule in paper
update model typing
update termination proof for dynamic typing
update notes
update static/proof subtyping in lean
update lfp elim rules in lean
update diff intro rule in lean
udpate proof subtyping rules in lean

\newcommand{\multi}[1]{\widebar{\ #1\ }}
\newcommand{\hastype}{:}
\newcommand{\pattype}{\ \lozenge\ }
\newcommand{\liftfun}{:}

@@

\begin{definition}
  \label{def:model_core_typing}
  \emph{model core typing}
  \hfill
  \small
  \boxed{\delta \VDash e \hastype \tau}
  \nopad
  \begin{mathpar}
    \inferrule { 
      \alpha \wr^+ \tau
      \\
      \exists n \qua e \hastype \tau[\alpha \slash \J{BOT}]^n 
    } {
      \delta \VDash e \hastype \J{LFP[}\alpha\J{]}\tau 
    } 
  \end{mathpar}
\end{definition}

\begin{definition}
  \label{def:model_finite_typing}
  \emph{model finite typing}
  \hfill
  \small
  \boxed{e \hastype \tau}
  \nopad
  \begin{mathpar}
  \end{mathpar}
\end{definition}

\begin{definition}
  \label{def:model_typing}
  \emph{model typing}
  \hfill
  \small
  \boxed{\delta \satisfies e \hastype \tau}
  \nopad
  \begin{mathpar}
    % \inferrule { 
    %   e \rightsquigarrow e'
    %   \\
    %   \delta \satisfies e' \hastype \tau 
    % } {
    %   \delta \satisfies e \hastype \tau 
    % }

    % \inferrule { 
    %   \alpha \slash \tau \in \delta
    %   \\
    %   \delta \satisfies e \hastype \tau
    % } {
    %   \delta \satisfies e \hastype \alpha 
    % }

    % \inferrule { 
    %   \TODO{alt def}
    %   \\\\
    %   \forall \tau' \qua  
    %   \delta \satisfies \tau[\alpha\slash\tau'] \subtypes \tau'
    %   \implies
    %   \delta \satisfies e \hastype \tau'
    % } {
    %   \delta \satisfies e \hastype \J{LFP[} \alpha \J{]} \tau
    % }

@@

@@

    \inferrule {
      \Theta, \Delta \Vvdash \Omega
      \\
      \Theta, \Delta \given \Omega \given \Theta', \Delta'
      \\
      \Theta' \cup \Theta_r, \Delta' \entails \tau_l \subtypes \tau_r \given \Theta'', \Delta'' 
    } {
      \Theta, \Delta \entails \tau_l \subtypes \J{ALL[}\Theta_r\J{]}\Omega \J{:} \tau_r \given \Theta'', \Delta''
    }
  \end{mathpar}
\end{definition}

@@

    \inferrule {
      \alpha \in \Theta 
      \\
      \tau' \J{<:} \alpha \in \Delta
      \\\\
      (\forall \alpha' \qua \tau' = \alpha' \implies \alpha' \in \Theta) 
      \\
      \Theta, \Delta \entails \tau \subtypes \tau' \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails 
      \tau \subtypes \alpha 
      \given \Theta', \Delta' 
    }
  \end{mathpar}
\end{definition}

@@

    \inferrule {
      \alpha \in \Theta 
      \\
      \alpha \J{<:} \tau' \in \Delta
      \\\\
      (\forall \alpha' \qua \tau' = \alpha' \implies \alpha' \in \Theta) 
      \\
      \Theta, \Delta \entails \tau' \subtypes \tau \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails 
      \alpha \subtypes \tau
      \given \Theta', \Delta'
    }
  \end{mathpar}
\end{definition}

@@

\begin{definition} 
  \label{def:proof_subtyping_lfp_elimination}
  \emph{proof subtyping (least fixed point elimination)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \J{LFP[}\alpha\J{]}\tau_l \subtypes \tau_r \given \Theta', \Delta'}
  \nopad
  \begin{mathpar}
    \inferrule {
      \Vdash \J{LFP[}\alpha\J{]}\tau_l \subtypes l \mapsto \tau_r' 
      \\
      \Theta, \Delta \entails \tau_r' \subtypes \tau_r \given \Theta', \Delta' 
    } {
      \Theta, \Delta \entails 
      \J{LFP[}\alpha\J{]}\tau_l \subtypes \J{<} l \J{>} \tau_r 
      \given \Theta', \Delta' 
    }

@@

    \inferrule {
      \tau_l[\alpha \slash \tau_r]^+ = \tau^\dagger
      \\
      \Theta, \Delta \entails 
      \tau^\dagger \subtypes \tau_r \given 
      \Theta', \Delta'
      \\
      \alpha \not\in \textbf{ftv}(\tau^\dagger)
      \\
      \alpha \not\in \textbf{ftv}(\tau_r)
    } {
      \Theta, \Delta \entails 
      \J{LFP[}\alpha\J{]}\tau_l \subtypes \tau_r 
      \given \Theta', \Delta' 
    }
  \end{mathpar}
\end{definition}

@@

\begin{definition} 
  \label{def:proof_subtyping_difference_intro}
  \emph{proof subtyping (difference introduction)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \tau_l \subtypes \tau_r \typdiff \eta \given \Theta', \Delta'}
  \nopad
  \begin{mathpar}
    \inferrule {
      \TODO{Update proof, deriving that lower does not intersect with subtraction}
      \\\\
      \Theta, \Delta \Vvdash \eta
      \\
      \nexists \Theta', \Delta' \qua  
      \Theta, \Delta \entails 
      \tau_l \subtypes \eta \given \Theta', \Delta'
      \\\\
      \Theta, \Delta \Vvdash \eta \subtypes  \tau_l
      \\
      \nexists \Theta', \Delta' \qua  
      \Theta, \Delta \entails 
      \eta \subtypes \tau_l \given \Theta', \Delta'
      \\\\
      \Theta, \Delta \entails 
      \tau_l \subtypes \tau_r \given \Theta', \Delta' 
    } {
      \Theta, \Delta \entails 
      \tau_l \subtypes \tau_r \typdiff \eta
      \given \Theta', \Delta'
    }
  \end{mathpar}
\end{definition}

@@

\begin{definition} 
  \label{def:proof_subtyping_lfp_intro}
  \emph{proof subtyping (least fixed point introduction)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \tau_l \subtypes \J{LFP[} \alpha \J{]}\tau_r \given \Theta', \Delta'}
  \nopad
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta \entails \tau_l \circlearrowleft \J{LFP[} \alpha \J{]} \tau_r
      \\
      \Theta, \Delta \entails
      \tau_l \subtypes \tau_r[\alpha \slash \J{LFP[} \alpha \J{]} \tau_r]
      \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails
      \tau_l \subtypes \J{LFP[} \alpha \J{]}\tau_r \given \Theta', \Delta'
    }

    \inferrule {
      \alpha \Vdash \tau_r \supertypes \tau_r'
      \\
      \Theta, \Delta \entails
      \tau_l \subtypes \tau_r' \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails
      \tau_l \subtypes \J{LFP[} \alpha \J{]}\tau_r \given \Theta', \Delta'
    }

@@

\noindent
For least fixed point introduction \ms{
\tau_l \subtypes \J{LFP[} \alpha \J{]}\tau_r
}, the solver has a couple options.
One option is to check for a syntactic guarantee that the least fixed point is inflatable without diverging 
\ms{
\tau_l \circlearrowleft \J{LFP[} \alpha \J{]} \tau_r
} and then solve the inflated introduction.
Alternatively, it can strengthen the least fixed point
with union drop \ms{\alpha \Vdash \tau_r \supertypes \tau_r'},
which drops any parts of the union containing the self reference.

\begin{definition} 
  \label{def:proof_subtyping_diff_elimination}
  \emph{proof subtyping (difference elimination)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \tau_l \typdiff \eta \subtypes \tau_r \given \Theta', \Delta'}
  \nopad
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta \entails 
      \tau_l \subtypes \eta \J{|} \tau_r \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails 
      \tau_l \typdiff \eta \subtypes \tau_r \given \Theta', \Delta'
    }
  \end{mathpar}
\end{definition}

@@

    \inferrule {
      \Vdash \tau_{l}\J{\&}\tau_{r} \subtypes \tau_x \rightarrow \tau_y 
      \\\\
      \Theta, \Delta \entails \tau_x \J{->} \tau_y  \subtypes \tau_a \J{->} \tau_b \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails \tau_{l}\J{\&}\tau_{r}  \subtypes \tau_a \J{->} \tau_b \given \Theta', \Delta'
    }

>>>>

\newcommand{\multi}[1]{\widebar{\ #1\ }}
\newcommand{\hastype}{:}
\newcommand{\pattype}{\ \lozenge\ }
\newcommand{\restricted}{\ \blacklozenge\ }
\newcommand{\liftfun}{:}

@@

% \begin{definition}
%   \label{def:model_progress_typing}
%   \emph{model progressive typing}
%   \hfill
%   \small
%   \boxed{\delta \satisfies e \hastype^* \tau}
%   \nopad
%   \begin{mathpar}
%     \inferrule { 
%       \downharpoonleft e
%       \up
%       \delta \satisfies e \hastype \tau 
%     } {
%       \delta \satisfies e \hastype^* \tau 
%     }

%     \inferrule { 
%       e \rightsquigarrow e'
%       \\
%       \delta \satisfies e' \hastype^* \tau 
%     } {
%       \delta \satisfies e \hastype^* \tau 
%     } 
%   \end{mathpar}

% \end{definition}

\begin{definition}
  \label{def:model_typing}
  \emph{model typing}
  \hfill
  \small
  \boxed{\delta \satisfies e \hastype \tau}
  \nopad
  \[
  \begin{array}[t]{rclcl}
      \delta \satisfies e \hastype \J{@}
      &\iff& 
      e \rightsquigarrow^* \J{@}

      \delta \satisfies e \hastype \J{<} l \J{>} \tau
      &\iff& 
      \delta \satisfies e\J{.}l \hastype \tau

      \delta \satisfies e \hastype \tau_l \J{->} \tau_r
      &\iff& 
      \forall e' \qua
      \delta \satisfies e' \hastype \tau_l 
      \implies
      \delta \satisfies e\J{(}e'\J{)} \hastype \tau_r

      \delta \satisfies e \hastype \tau_l \J{|} \tau_r 
      &\iff& 
      \delta \satisfies e \hastype \tau_l
      \down
      \delta \satisfies e \hastype \tau_r

      \delta \satisfies e \hastype \tau_l \J{\&} \tau_r 
      &\iff& 
      \delta \satisfies e \hastype \tau_l
      \up
      \delta \satisfies e \hastype \tau_r

      \delta \satisfies e \hastype \tau_l \typdiff \tau_r
      &\iff& 
      \delta \satisfies e \hastype \tau_l
      \up
      \delta \not\satisfies e \hastype \tau_r

      \delta \satisfies e \hastype \J{EXI[}\Theta\J{]}\Delta \J{:} \tau
      &\iff& 
      \exists \delta' \qua
      \textbf{dom}(\delta') \subseteq \Theta
      \up
      \delta \oplus \delta' \satisfies \Delta 
      \up
      \delta \oplus \delta' \satisfies e \hastype \tau

      \delta \satisfies e \hastype \J{ALL[}\Theta\J{]}\Delta \J{:} \tau
      &\iff& 
      \forall \delta' \qua
      \textbf{dom}(\delta') \subseteq \Theta
      \implies
      \delta \oplus \delta' \satisfies \Delta 
      \implies
      \delta \oplus \delta' \satisfies e \hastype \tau

      \delta \satisfies e \hastype \J{LFP[}\alpha\J{]}\tau 
      &\iff& 
      \alpha \wr^+ \tau
      \up
      \exists n \qua \VDash e \hastype (\tau[\alpha \slash \J{BOT}]^n)[\delta]

      \delta \satisfies e \hastype \alpha 
      &\iff& 
      \alpha \slash \tau \in \delta
      \up
      \VDash e \hastype \tau
  \end{array}
  \]
\end{definition}

\begin{definition}
  \label{def:model_finite_typing}
  \emph{model finite typing}
  \hfill
  \small
  \boxed{\VDash e \hastype \tau}
  \nopad
  \[
  \begin{array}[t]{rclcl}

      \VDash e \hastype \J{@}
      &\iff& 
      e \rightsquigarrow^* \J{@}

      \VDash e \hastype \J{<} l \J{>} \tau
      &\iff& 
      \VDash e\J{.}l \hastype \tau

      \VDash e \hastype \tau_l \J{->} \tau_r
      &\iff& 
      \forall e' \qua
      \VDash e' \hastype \tau_l 
      \implies
      \VDash e\J{(}e'\J{)} \hastype \tau_r

      \VDash e \hastype \tau_l \J{|} \tau_r 
      &\iff& 
      \VDash e \hastype \tau_l
      \down
      \VDash e \hastype \tau_r

      \VDash e \hastype \tau_l \J{\&} \tau_r 
      &\iff& 
      \VDash e \hastype \tau_l
      \up
      \VDash e \hastype \tau_r

      \VDash e \hastype \tau_l \typdiff \tau_r
      &\iff& 
      \VDash e \hastype \tau_l
      \up
      \not\VDash e \hastype \tau_r
  \end{array}
  \]
\end{definition}

\TODO{note that the semantics allows progam to get stuck under vacuously inhabited types}

% \begin{definition}
%   \label{def:model_scratch_typing}
%   \emph{model scratch typing}
%   \hfill
%   \small
%   \boxed{\delta \satisfies e \hastype \tau}
%   \nopad
%   \begin{mathpar}
    % \inferrule { 
    %   \TODO{alt def}
    %   \\\\
    %   \forall \tau' \qua  
    %   \delta \satisfies \tau[\alpha\slash\tau'] \subtypes \tau'
    %   \implies
    %   \delta \satisfies e \hastype \tau'
    % } {
    %   \delta \satisfies e \hastype \J{LFP[} \alpha \J{]} \tau
    % }

@@

%   \end{mathpar}
% \end{definition}

@@

    \inferrule {
      \Theta, \Delta \Vvdash \Omega
      \\
      \Theta, \Delta \given \Omega \entails \Theta', \Delta'
      \\
      \Theta' \cup \Theta_r, \Delta' \entails \tau_l \subtypes \tau_r \given \Theta'', \Delta'' 
    } {
      \Theta, \Delta \entails \tau_l \subtypes \J{ALL[}\Theta_r\J{]}\Omega \J{:} \tau_r \given \Theta'', \Delta''
    }
  \end{mathpar}
\end{definition}

@@

    \inferrule {
      \alpha \in \Theta 
      \\
      \tau' \J{<:} \alpha \in \Delta
      \\\\
      (\forall \alpha' \qua \alpha' = \tau' \implies \alpha' \in \Theta) 
      \\
      \Theta, \Delta \entails \tau \subtypes \tau' \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails 
      \tau \subtypes \alpha 
      \given \Theta', \Delta' 
    }
  \end{mathpar}
\end{definition}

@@

    \inferrule {
      \alpha \in \Theta 
      \\
      \alpha \J{<:} \tau' \in \Delta
      \\\\
      (\forall \alpha' \qua \alpha' = \tau' \implies \alpha' \in \Theta) 
      \\
      \Theta, \Delta \entails \tau' \subtypes \tau \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails 
      \alpha \subtypes \tau
      \given \Theta', \Delta'
    }
  \end{mathpar}
\end{definition}

@@

\begin{definition} 
  \label{def:proof_subtyping_lfp_elimination}
  \emph{proof subtyping (least fixed point elimination)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \J{LFP[}\alpha\J{]}\tau_l \subtypes \tau_r \given \Theta', \Delta'}
  \nopad
  \begin{mathpar}
    \inferrule {
      \Vdash \J{LFP[}\alpha\J{]}\tau_l \subtypes l \mapsto \tau_f 
      \\
      \Theta, \Delta \entails \tau_f \subtypes \tau_r \given \Theta', \Delta' 
    } {
      \Theta, \Delta \entails 
      \J{LFP[}\alpha\J{]}\tau_l \subtypes \J{<} l \J{>} \tau_r 
      \given \Theta', \Delta' 
    }

@@

    \inferrule {
      \alpha \wr^+ \tau_l
      \\
      \Theta, \Delta \entails 
      \tau_l[\alpha \slash \tau_r] \subtypes \tau_r \given 
      \Theta', \Delta'
    } {
      \Theta, \Delta \entails 
      \J{LFP[}\alpha\J{]}\tau_l \subtypes \tau_r 
      \given \Theta', \Delta' 
    }
  \end{mathpar}
\end{definition}

@@

\begin{definition} 
  \label{def:proof_subtyping_difference_intro}
  \emph{proof subtyping (difference introduction)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \tau \subtypes \tau_l \typdiff \tau_r \given \Theta', \Delta'}
  \nopad
  \begin{mathpar}
    \inferrule {
      \TODO{Update proof, deriving that lower does not intersect with subtraction}
      \\\\
      \TODO{in lean, prove that subtyping is equivalent to decidable definition under restricted subtyping form}
      \\\\
      \epsilon \pattype \tau_r
      \\
      \Theta, \Delta \restricted \tau_r \subtypes  \tau
      \\\\
      \Theta, \Delta \not\Vvdash \tau \subtypes \tau_r
      \\
      \Theta, \Delta \not\Vvdash \tau_r \subtypes \tau
      \\
      \Theta, \Delta \entails \tau \subtypes \tau_l \given \Theta', \Delta' 
    } {
      \Theta, \Delta \entails 
      \tau \subtypes \tau_l \typdiff \tau_r
      \given \Theta', \Delta'
    }
  \end{mathpar}
\end{definition}

@@

\begin{definition} 
  \label{def:proof_subtyping_lfp_intro}
  \emph{proof subtyping (least fixed point introduction)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \tau_l \subtypes \J{LFP[} \alpha \J{]}\tau_r \given \Theta', \Delta'}
  \nopad
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta \circlearrowleft \tau_l \subtypes \J{LFP[} \alpha \J{]} \tau_r
      \\
      \Theta, \Delta \entails
      \tau_l \subtypes \tau_r[\alpha \slash \J{LFP[} \alpha \J{]} \tau_r]
      \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails
      \tau_l \subtypes \J{LFP[} \alpha \J{]}\tau_r \given \Theta', \Delta'
    }

    \inferrule {
      \Theta, \Delta \entails
      \tau_l \subtypes \textbf{drop}(\alpha \entails \tau_r) \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails
      \tau_l \subtypes \J{LFP[} \alpha \J{]}\tau_r \given \Theta', \Delta'
    }

@@

\noindent
For least fixed point introduction \ms{
\tau_l \subtypes \J{LFP[} \alpha \J{]}\tau_r
}, the solver has a couple options.
One option is to check for a syntactic guarantee that the least fixed point is inflatable without diverging 
\ms{
... \circlearrowleft \tau_l \subtypes \J{LFP[} \alpha \J{]} \tau_r
} and then solve the inflated introduction.
Alternatively, it can strengthen the least fixed point
with union drop \ms{\alpha \Vdash \tau_r \supertypes \tau_r'},
which drops any parts of the union containing the self reference.

\begin{definition} 
  \label{def:proof_subtyping_diff_elimination}
  \emph{proof subtyping (difference elimination)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \tau_l \typdiff \tau_r \subtypes \tau \given \Theta', \Delta'}
  \nopad
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta \entails 
      \tau_l \subtypes \tau_r \J{|} \tau \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails 
      \tau_l \typdiff \tau_r \subtypes \tau \given \Theta', \Delta'
    }
  \end{mathpar}
\end{definition}

@@

    \inferrule {
      \Theta, \Delta \entails \textbf{merge}(\tau_{l}\J{\&}\tau_{r}) \subtypes \tau_a \J{->} \tau_b \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails \tau_{l}\J{\&}\tau_{r}  \subtypes \tau_a \J{->} \tau_b \given \Theta', \Delta'
    }
